---
title: "Causal Diagrams & Model Planning"
author: ""
date: "June 2025"
output:
  xaringan::moon_reader:
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '4:3'
---

```{r, setup-lm1, include=FALSE}
library(ggformula)
library(tidyverse)
library(dagitty)
library(CalvinBayes)

knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 7, 
                      fig.height = 3,
                      tidy = FALSE,
                      fig.align = 'center', 
                      message = FALSE, 
                      warning = FALSE,
                      error = TRUE,
                      out.width = '60%', 
                      dpi = 300)
theme_set(theme_minimal(base_size = 22))

bonobos <- read_csv(file='http://sldr.netlify.app/data/bonobo_faces.csv')
# add one missing for demo purposes
bonobos[28, 'normDS'] <- NA

m3_2q2c <- lm(AssR ~ fWHR + normDS + Sex + Group, 
           data = bonobos)
```



# Causal Diagrams
### There's more to planning than just p < n/15!

```{r, simple-causal-diagram, echo = FALSE, out.width='85%'}
# create the DAG object
causal_diagram_1 <- dagitty("dag{
  X -> Y;
}")
# plot it
gg_dag(causal_diagram_1, size = 16)
```

---
## "Causal"? 
### Can we actually make *causal conclusions*?

.small[
- If detect association, can conclude \_\_\_ *causes* \_\_\_ IFF...
<br>
<br>
- Otherwise:
<br>
<br>
- So why use "causal" diagrams to plan models?
  - Transparency about assumptions
  - Clear, consistent way to choose predictors
  ]

---

## H.E.A.L.T.H. Camp Example
### Response: Test Score

.pull-left[
- Age
- Sex 
- Grade in school
- Neighborhood
- Race & Ethnicity
]

.pull-right[
- Income
- Interest in health career
- Timepoint (before/after camp)
]
---
# Confounder

---
# Precision Covariate

---
# Mediator

---
# Moderator or Modifier
## Also known as: *Interaction*
---
# Collider

---
# Can get complex
## M-Bias

```{r, m-bias-diagram, echo = FALSE, out.width='85%'}
# create the DAG object
causal_diagram_2 <- dagitty("dag{
  X -> Y;
  A -> X;
  A -> D;
  B -> Y;
  B -> D;
}")
# plot it
gg_dag(causal_diagram_2, size = 16, highlight = c('X', 'Y'))
```

.small[
*We are using a some simple principles but there could be much more to learn!*
]
.smaller[
*Resource: Guide to Causal Inference <https://doi.org/10.1098/rspb.2020.2815>*
]
---
# Look back
## which variables should be predictors
## to model HEALTH Camp test score?

<br>
# score ~ 
---

# Your Summary
## Linear modeling step-by-step:
