---
title: "Count Data and Models"
subtitle: "June 2025"
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(viridis)
library(sugrrants)
library(lubridate)
library(ggformula)
```

### What does it mean to be count data?
- Constrained to be positive
  - Zero or greater

--
- Discrete
  - Whole number counts

--
- Rate
  - Counts per unit


---
### Lizard Data Set

```{r loaddata,echo=FALSE, message=FALSE, warning=FALSE}
lizard <- read_csv("lizard.csv")
```

.question[
Does vegetation or temperature explain the number of sand lizards found at a site?
]

--

Variables:

lizard ~ veg + temp + visit

- `lizard`: the number of sand lizards found at one of 200 sites
- `veg`: A measure of vegetation between -1 and 1, where 0 is the average vegetation density  
- `temp`: A measure of temperature between -1 and 1, where 0 is the average temperature
- `visit`: Takes the value 1, 2 or 3 depending on which visit to the site is being recorded

Data courtesy of Marc Kery

---
### Visualize
```{r hist, fig.align='center',fig.width=6,fig.height=6}
ggplot(lizard,aes(x=lizard))+
  geom_histogram(bins=10) +
  labs(x="Number of Lizards",
       y="Count")
```

---
### Lizards and Vegetation
.small[
```{r plot,fig.align='center',fig.width=6,fig.height=6}
ggplot(lizard,aes(y=lizard,x=veg))+
  geom_point(size=3)+
  labs(title="Lizards by Vegetation",
       x = "Vegeitation",
       y = "Number of Lizards")
```
]

---
### Bad Idea
```{r badplot,fig.align='center',fig.width=6.5,fig.height=6.5,echo=FALSE, message=FALSE}
ggplot(lizard,aes(y=lizard,x=veg))+
  geom_point(size=3) + 
  labs(title="Lizards by Vegitation",
       x = "Vegitation",
       y = "Number of Lizards")+
  geom_smooth(method="lm",se=F)
```

---
### Why bad?

---
### Check Formally

.pull-left[
- Independence
```{r badacf, fig.align='center',fig.width=6,fig.height=4,echo=FALSE}
liz.lm    <- lm(lizard~veg,data=lizard)
resid.lm  <- residuals(liz.lm)
fitted.lm <- fitted(liz.lm)
acf(resid.lm)
```
]
.pull-right[
- Error is constant
```{r badvar, fig.align='center',fig.width=6,fig.height=4,echo=FALSE}
gf_point(resid.lm ~ fitted.lm)
```
]

---
### Normality

.pull-left[
- Histogram
```{r badhis, echo=FALSE}
gf_histogram(~resid.lm, bins=15) 
```
]
.pull-right[
- QQ plot
```{r badqq,echo=FALSE}
gf_qq(~resid.lm) |>
  gf_qqline()
```
]


---
### Good Idea
.small[
```{r goodplot,fig.align='center',fig.width=6,fig.height=6,message=FALSE}
ggplot(lizard,aes(y=lizard,x=veg))+
  geom_point(size=3) + 
  labs(title="Lizards by Vegitation",
       x = "Vegitation",
       y = "Number of Lizards")+
  geom_smooth(method="glm",method.args = list(family = "poisson"),se=F)
```
]


---
### How do we get there?

--
- Generalized linear model (GLM)

--
- Relax some assumptions
  - Linear on the scale of the linear predictor
  - Residuals are appropriate for the chosen distribution

--
- Still need independence

---
### Poisson GLM
- Response variable
$$lizard_{i} \sim Pois(\lambda_{i})$$

--
- Explanatory variable
$$g(\lambda_{i}) = \eta_{i} = \beta_{0} + \beta_{1}veg_{i}$$

--
- Link function
$$g(\lambda_{i}) = \log(\lambda_{i})=\eta_{i} = \beta_{0} + \beta_{1}veg_{i}$$

--
- Invertible
$$\lambda_{i} = g^{-1}(\eta_{i}) = \exp(\beta_{0} + \beta_{1}veg_{i})$$

---
### Model Fitting
```{r loglin}
pois.mod <- glm(lizard~veg+temp+as.factor(visit),family=poisson(link="log"),data=lizard)
summary(pois.mod)
```

---
### Assumptions

.pull-left[
.small[
```{r assumptions}
resid <- residuals(pois.mod,type="response")
fit   <- fitted(pois.mod)
plot(fit,resid,main="Residual vs Fitted",xlab="Fitted",ylab="Residuals")
abline(h=0)
```
 ]
]
.pull-right[
.small[
```{r poisacf, fig.align='center'}
acf(resid)
```
 ]
]

---
### Right amount of variance?

- Use Pearson's residuals
  - Scaled by expected variance
  - Approximately constant variance
  - Standard deviation of 1

.small[
```{r pearsonresid, fig.align='center',fig.width=5,fig.height=5}
p.resid <- resid(pois.mod, type='pearson')
gf_point(p.resid ~ fit) |>
  gf_labs(x='Fitted Values', y='Pearson Residuals')
```
]

---
### Overdispersion

- More variance than expected
  - Poisson distribution: $\mu$ = $\sigma^2$
  
--
- Can check with the quasi-Poisson
.small[
```{r quasi}
quasi.mod <- glm(lizard~veg+temp+as.factor(visit),family=quasipoisson(),data=lizard)
summary(quasi.mod)
```
]

---
## How big is too big?
  - overdispersion < 1.5, generally okay
  - overdispersion > 15, big problems

--
- Alternative is the negative-binomial
  - Variance can be greater than the mean
  - Still using the log link

.small[  
```{r negbin,message=FALSE}
library(MASS)
nb.mod <- glm.nb(lizard~veg+temp+as.factor(visit), data=lizard)
summary(nb.mod)
```
]

---
### Which to use?

- Judgement call

--
- Can compare

```{r compare, message=FALSE}
library(lmtest)
lrtest(pois.mod,nb.mod)
```

---
### Assessment

.question[What might be a good why to assess model fit?]

--
- $R^{2}$ works best for simple linear regression
  - Are some other related options, won't cover here

--
- Observed versus predicted
```{r obsvpred,fig.align='center',fig.width=5,fig.height=5,echo=FALSE}
plot(fitted(nb.mod),lizard$lizard,main="Observed vs Predicted",ylab="Observed Lizards",xlab="Predicted Lizards")
abline(0,1)
```

---
### Intepretation

.question[How do we interpret the intercept in this case?]

--
- Zero is the average in our data
  - Number of lizards for average vegetation and temperature
  - $\exp(0.434)=1.54$

--
- Start with the sign of the slopes
  - Decrease with increasing vegetation
  - Increase with increasing temperature
  - No effect of visit (slopes not statistically different from zero)

---
### More Interpretation

.question[What is the difference between the scale of the linear predictor and scale of the response?]

--
- Exponentiate the slopes
  - Percentage change in response variables for a unit increase in the explanatory variable
  - $\exp(-0.569)=0.566$
  - $\exp(1.101)=3.01$

--
- What is a unit?

--
- -1 to 0
- 0 to 1

--
- Scale by raising to the power
  - $\exp(-0.569)^{0.1}=0.95$
  - $\exp(1.101)^{0.1}=1.12$


