---
title: "AIC and ANOVA"
subtitle: "June 2025"
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(viridis)
library(sugrrants)
library(lubridate)
library(ggformula)
library(MASS)
```

### What to do with our model set?
- We were excellent planners
  - Developed clear hypotheses
  - Model set checked and ready to go

--
- Now what?

--
- Choose the "best" model
  - AIC
  - ANOVA


---
### Crime

```{r, message=FALSE, echo=FALSE}
crime <- read_csv("crimedata.csv")
```

.question[
Does education help explain the rate of violent crime?
]

--

Variables:
- `Total`: the total number of crimes per million
- `Violent`: the number of violent crimes per 100000  
- `Police`: Police funding in dollars per resident per year
- `Highschool`: Percent of adults with 4 years of high school
- `Deliquent`: Percent of 16-19 years olds not in high school
- `College`: Percent of 18-24 year olds in college
- `CollegeAdults`: Percent of adults with at least 4 years of college

---
### What's our plan?

--
$n/15 = 50/15 = 3.33$
 - Not going to use all the predictors
 - Construct hypotheses
 
--
- Control all hypotheses for the amount of money spent on policing

--
- Hypothesis #1:
  - Low education levels are a the driver of violent crime
.small[
```{r hyp1}
crime1 <- glm(Violent~Police+HighSchool+Deliquent,family = poisson(), data=crime)
```
]

- Hypothesis #2:
  - More college educated adults means less violent crime
.small[
```{r hyp2}
crime2 <- glm(Violent~Police+CollegeAdults,family=poisson(),data=crime)
```
]

---
### Coding Time

This is a continuation of our planning activity
 - Make additions and modifications to the previous Quarto file
 - Keep the same initial model
 - Develop, fit and assess a second model hypothesis
 
---
### Choices, choices

- Akaike’s Information Criterion (AIC)
  - Penalizes for the number of estimated parameters
  $$-2log-likelihood + 2p$$
  - Roughly the same sample size between models being compared

--
- Measures the information lost
  - Goal is the least amount of lost information

--
- Lowest AIC = "best" model
  - Marked preference when the difference between AIC values is greater than 2
  - Subjective

--
- Picks the "best" model in the set
  - Doesn't mean its any good
  - A better one could always exist
  
---
### Different Flavours of AIC

- Correct for small sample size, AICc
$$-2log-likelihood + 2p + 2\frac{p(p+1)}{(n-p-1)}$$

--
- Handle over-dispersed data, QAIC
$$QAIC = -2\frac{2}{VIF}log-likelihood + 2p$$
  - Can do QAICc too!

--
- Measures relative, not absolute, fit

---
### Finding AIC
`AIC` in `R`

```{r aic}
AIC(crime1,crime2)
```

--
The model with the covariates related to high school education performs better

--
## WARNING!
AIC is only defined up to an arbitrary constant
 - Different constants for different purposes
 - Take care when comparing fits of different classes

---
### Coding Time
Use AIC to select between the two models you defined earlier

---
### Analysis of Variance (ANOVA)

.question[Is one of our models doing a better job explaining the variability in the data?]


--
Variability is divided into two parts:
 - Explained by the model - Regression sums of squares
 - Unexplained by the model - Residual sums of squares

---
### How's it work?

Models with more explanatory variables should explain more variability

--
- If true, keep the model with more covariates

--
- If the models explain a similar amount of variability, keep the model with fewer covariates  

--
$H_{0}:$ The two models are equivalent   
$H_{1}:$ The model with more covariates explains more variability  

---
### ANOVA

```{r aov}
anova(crime1, crime2)
```
--
Strong evidence against the null hypothesis

---
### Null Model

.question[Do we learn anything at al from doing regression?]


--
- ANOVA to compare full and null model
  - Full model: the one with explanatory variables
  - Null model: intercept only
  - The hypothesis test at the end of the `lm` output

--
```{r full}
crime.null <- glm(Violent~1, family=poisson(), data=crime)
anova(crime1,crime.null)
```

---
### Now you try

- Compare your two models using ANOVA
- Build a null model and compare to the chosen model using ANOVA
