---
title: "Predictions & Prediction Plots"
author: ""
date: "June 2025"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '4:3'
---

```{r, setup, include=FALSE}
library(tidyverse)
library(ggformula)
library(ggeffects)
library(glmmTMB)

knitr::opts_chunk$set(fig.width = 10, fig.height = 6, fig.align = "center")
theme_set(theme_minimal(base_size = 24))
```

class: huge center inverse subsection
<br>
# What Does It* All Mean?
<br>
### *Especially all those $\hat{\beta}$s...

---

# Task #1: 
### A. Write model equation
### B. In words, how does [outcome] depend on [predictor]?

```{r, echo = FALSE}
upf_data <- read_csv("https://sldr.netlify.app/data/upf_by_diet.csv",
                     show_col_types = FALSE)
upf_model <- lm(weight_change ~ diet + age + sex + first_diet,
                data = upf_data)
```

.smaller[
```{r, echo = FALSE}
summary(upf_model)
```
]

---
# Task #2
### A. Write model equation
### B. In words, how does [outcome] depend on [predictor]?

```{r, echo = FALSE}
cancer_data <- read_csv("https://sldr.netlify.app/data/LA_cancer_data.csv",
                        show_col_types = FALSE) |>
  mutate(log_population = log(annual_population),
         point_cancer_risk = point_cancer_risk * 1000,
         nonpoint_cancer_risk = nonpoint_cancer_risk * 1000)
cancer_model <- glmmTMB(cancer_cases ~ point_cancer_risk +
                          nonpoint_cancer_risk +
                          percent_poverty +
                          percent_smokers +
                          percent_construction + 
                          percent_manufacturing +
                          offset(log_population),
                        data = cancer_data,
                        family = nbinom2(link = "log"))
```

.smaller[
```{r, echo = FALSE, message = FALSE, warning = FALSE}
mosaic::msummary(cancer_model)
```
]

---
# Why Not This?

```{r, fig.width = 8.5, fig.height=4, warning = FALSE, message = FALSE}
gf_point(weight_change ~ age, 
         data = upf_data) |>
  gf_lm()
```

---
# Which to Present?
.small[
```{r, warning = FALSE}
coef(lm(weight_change ~ age,
           data = upf_data)) 
```


```{r, warning = FALSE}
coef(lm(weight_change ~ age + 
          diet + sex + 
          first_diet,
           data = upf_data)) 
```

]
---
## Adding "the line" to "the plot"

```{r, echo = FALSE, fig.width = 12, fig.height = 6}
gf_point(weight_change ~ age,
         alpha = 0.5,
         size = 4,
         data = upf_data) |>
  gf_line(preds ~ age, data = upf_data |>
            drop_na(age, diet, sex, first_diet) |>
            mutate(preds = predict(upf_model)),
          color = "darkred", alpha = 0.5,
          linewidth = 1.75) |>
  gf_point(preds ~ age, data = upf_data |>
            drop_na(age, diet, sex, first_diet) |>
            mutate(preds = predict(upf_model)),
          color = "darkred",
          shape = 17, size = 3)
```

---

class: huge center inverse section

<br>
# So...
# what could we do?

---
# Model Predictions...
### For *fake* data???


```{r}
library(ggeffects)
upf_predictions <- 
  predict_response(upf_model,
            terms = "age")
```


---
# Model Predictions...
### For *fake* data???

.smaller[
```{r}
upf_predictions
```
]

---
# See "fixed" value(s)

.smaller[
```{r}
predict_response(
  upf_model,
  terms = "age",
  condition = c(diet = "Ultra-processed"))
```
]

---
# Plot it!

```{r}
plot(upf_predictions)
```

---
## Small Customizations

.smaller[
```{r, echo = TRUE, eval = FALSE}
plot(upf_predictions) |>
  gf_labs(x = "Age", y = "Weight Change (kg)",
          title = "UPF weight gain roughly constant with age",
          subtitle = "(model-predicted change for males on UPF diet)")
```
]

```{r, echo = FALSE, eval = TRUE}
plot(upf_predictions, use_theme = FALSE) |>
  gf_labs(x = "Age", y = "Weight Change (kg)",
          title = "UPF weight gain roughly constant with age",
          subtitle = "(model-predicted change for males on UPF diet)")
```
---
# Categorical Too
.small[
```{r, fig.width = 10, fig.height = 5}
predict_response(upf_model,
                 terms = "diet",
                 verbose = FALSE) |>
  plot(use_theme = FALSE)
```
]


---
## When this is *really* useful: GLMs

.smaller[
```{r, echo = TRUE, eval = FALSE}
predict_response(cancer_model,
                 terms = "percent_smokers",
                 condition = c(log_population = log(1000)),
                 verbose = FALSE)|>
  plot() |>
  gf_labs(x = "Percent Smokers",
          y = "Cancer Cases (per 1000 residents)")
```
]

```{r, echo = FALSE, eval = TRUE}
predict_response(cancer_model,
                 terms = "percent_smokers",
                 condition = c(log_population = log(1000)),
                 verbose = FALSE)|>
  plot(use_theme = FALSE) |>
  gf_labs(x = "Percent Smokers",
          y = "Cancer Cases (per 1000 residents)")
```

---

## Your Turn

## 1. Choose a model you fit before
## 2. Make a prediction plot! 

.small[
- Note the *fixed values* used. 
- What do you learn from the plot?
]

## 3. What happens if you use >1 `terms`? 

.small[ 
- Code will look like: `predict_response(..., terms = c("age", "diet"))`
]

## 4. What happens if you swap the order of the `terms` being plotted?