---
title: "Interaction Terms"
subtitle: "June 2025"
format: 
  revealjs:
    theme: solarized
editor: visual
---

## Interaction Terms

-   Relationship between the response and predictor varies with the value of another predictor

. . .

-   Hierarchy principle: main effects needed when interaction effect is present
    -   Main effect: the explanatory variable on its own
    -   Interaction effect: the product of two explanatory variables

## Case Study

Is there a relationship between economic growth and energy consumption?

 - Data from Goshrey et al. (2018)
 
. . . 

-   Variables
    -   `year` Calendar year
    -   `GDP` US gross domestic product
    -   `EC` Energy consumption in billion BTU
    -   `O_P` Oil price in dollars
    -   `SPE` Public spending in dollars 

## The Model

 - Ideally, control for spending and oil price
 - Energy consumption and oil price my interact to influence economic growth
 - Greater volatility in oil prices after 1972, may interact with energy consumption

## Creating a variable

```{r,echo=FALSE, warning=FALSE,message=FALSE}
library(tidyverse)
library(scales)
require(ggformula)
enrg <- read_csv("Data_Reassesing.csv")
```

```{r, echo=TRUE}
enrg <- enrg |>
        mutate(V = ifelse(year < 1973,0,1))
```

```{r}
knitr::kable(enrg)
```

## Regression Model: Categorical Interaction Term

$$GDP_{i} = \beta_{0} + \beta_{1}EC_{i} + \beta_{2}V_{i} + \beta_{3}EC_{i}*V_{i} + \epsilon_{i}$$ - $V_{i}$ is an indicator variable\
- Before 1972 has a different slope then after 1972

## When $V_{i} = 1$

$$GDP_{i} = \beta_{0} + \beta_{1}EC_{i} + \beta_{2}V_{i} + \beta_{3}EC_{i}*1 + \epsilon_{i}$$


$$GDP_{i} = \beta_{0} + (\beta_{1}EC_{i}+\beta_{3}EC_{i}) + \beta_{2}V_{i} + \epsilon_{i}$$


$$GDP_{i} = \beta_{0} + (\beta_{1}+\beta_{3})EC_{i} + \beta_{2}V_{i} + \epsilon_{i}$$

## When $V_{i} = 0$

$$GDP_{i} = \beta_{0} + \beta_{1}EC_{i} + \beta_{2}V_{i} + \beta_{3}EC_{i}*0 + \epsilon_{i}$$


$$GDP_{i} = \beta_{0} + \beta_{1}EC_{i} + \beta_{2}V_{i} + \epsilon_{i}$$

## Fitting the Model in `R`

```{r, echo=TRUE}
enrg.int <- lm(GDP~EC*V,data=enrg)
summary(enrg.int)
```

## Visualizing the Interaction: Code

```{r,warning=FALSE,message=FALSE,echo=TRUE, eval=FALSE}
ggplot(enrg,aes(EC,GDP))+
  geom_point() +
  facet_wrap(vars(V)) +
  geom_smooth(method="lm",se=FALSE) +
  labs(title="GDP and Energy Consumption",
       x = "Energy Consumption",
       y = "GDP") +
  scale_x_continuous(labels = label_comma()) +
  scale_y_continuous(labels = label_comma())

```

## Visualizing the Code: Plots
```{r,warning=FALSE,message=FALSE, echo=FALSE}
ggplot(enrg,aes(EC,GDP))+
  geom_point() +
  facet_wrap(vars(V)) +
  geom_smooth(method="lm",se=FALSE) +
  labs(title="GDP and Energy Consumption",
       x = "Energy Consumption",
       y = "GDP") +
  scale_x_continuous(labels = label_comma()) +
  scale_y_continuous(labels = label_comma())

```

## Coding Time!

Using the data from the most recent practical, fit a model with a categorical interaction term.  

If you are feeling ambitious, plot it!

## Regression Model: Quantitative Interaction Term

$$GDP_{i} = \beta_{0} + \beta_{1}EC_{i} + \beta_{2}OP_{i} + \beta_{3}EC_{i}*OP_{i} + \epsilon_{i}$$ - No simplification\
- How do we interpret?

## Visualize: No Interaction

```{r, echo=FALSE}
library(plot3D)
# set the x, y, and z variables
x <- enrg$O_P
y <- enrg$EC
z <- enrg$GDP

# Compute the linear regression 
fit <- lm(z ~ x + y)

# create a grid from the x and y values (min to max) and predict values for every point
# this will become the regression plane
grid.lines = 40
x.pred <- seq(min(x), max(x), length.out = grid.lines)
y.pred <- seq(min(y), max(y), length.out = grid.lines)
xy <- expand.grid( x = x.pred, y = y.pred)
z.pred <- matrix(predict(fit, newdata = xy), 
                 nrow = grid.lines, ncol = grid.lines)

# create the fitted points for droplines to the surface
fitpoints <- predict(fit)

# scatter plot with regression plane
scatter3D(x, y, z, pch = 19, cex = 1,colvar = NULL, col="red", 
          theta = 40, phi = 0, bty="b",
          xlab = "Oil Price", ylab = "Energy", zlab = "GDP",  
          surf = list(x = x.pred, y = y.pred, z = z.pred,  
                      facets = TRUE, fit = fitpoints, col=ramp.col (col = c("dodgerblue3","seagreen2"), n = 300, alpha=0.7), border="black"), main = "")

```

## Fitting the Model in `R`

```{r, echo=TRUE}
enrg.int.cont <- lm(GDP~EC*O_P,data=enrg)
summary(enrg.int.cont)
```

## Visualize: Interaction

```{r, echo=FALSE}
library(plot3D)
# set the x, y, and z variables
x <- enrg$O_P
y <- enrg$EC
z <- enrg$GDP

# Compute the linear regression 
fit <- lm(z ~ x*y)

# create a grid from the x and y values (min to max) and predict values for every point
# this will become the regression plane
grid.lines = 40
x.pred <- seq(min(x), max(x), length.out = grid.lines)
y.pred <- seq(min(y), max(y), length.out = grid.lines)
xy <- expand.grid( x = x.pred, y = y.pred)
z.pred <- matrix(predict(fit, newdata = xy), 
                 nrow = grid.lines, ncol = grid.lines)

# create the fitted points for droplines to the surface
fitpoints <- predict(fit)

# scatter plot with regression plane
scatter3D(x, y, z, pch = 19, cex = 1,colvar = NULL, col="red", 
          theta = 150, phi = 20, bty="b",
          xlab = "Oil Price", ylab = "Energy", zlab = "GDP",  
          surf = list(x = x.pred, y = y.pred, z = z.pred,  
                      facets = TRUE, fit = fitpoints, col=ramp.col (col = c("dodgerblue3","seagreen2"), n = 300, alpha=0.7), border="black"), main = "")

```

## Excluding Main Effects

```{r, echo=TRUE}
mod.only <- lm(GDP~EC:O_P,data=enrg)
summary(mod.only)

```

## Coding Time

Using the data from the most recent practical, fit a model with a quantitative interaction term.  

Using the data from the most recent practical, fit a model with only an interaction term


## Side Note

Bad residual vs fitted plot

```{r, echo=FALSE}
enrg <- enrg |>
  mutate(predsV = predict(enrg.int),
         residsV = resid(enrg.int),
         predsC = predict(enrg.int.cont),
         residsC = resid(enrg.int.cont))
```

::::: columns
::: column
```{r, fig.width=7, fig.hieght=6, echo=FALSE}
gf_point(residsV ~ predsV,
         data = enrg,
         xlab="Predicted Value",
         ylab="Residual",
         title="Interaction with Period")
```
:::

::: column
```{r, fig.width=7, fig.hieght=6,echo=FALSE}
gf_point(residsC ~ predsC,
         data = enrg,
         xlab="Predicted Value",
         ylab="Residual",
         title="Interaction with Oil Price")
```
:::
:::::