---
title: "The Many Flavours of Regression"
subtitle: "June 2025"
format: 
  revealjs:
    theme: moon
editor: visual
---

## Generalized Addative Models (GAMs)

Non-linear relationships

. . . 

What is meant by non-linear?

. . . 

-   Non-linear in the parameters

. . . 

Use a smooth function to link response and predictor
$$y_{i} = \beta_{0} + f(x_{i}) + \epsilon_{i}$$

## Case Study

What affects the body condition of fin whales in the North Atlantic?

-   Data collected from 1967-2010
-   Full analysis in Williams et al. 2013

. . .

Variables:

-   `LE` - length in meters
-   `m4` - a measure of body condition 
-   `NewStatus` - Pregnant (1) or not (0)
-   `SelP` - Zooplankton from Selvogsbanki


## Splines

Definitions:

-   A thin wood or metal strip used in building construction
-   A numeric function that is piecewise-defined by polynomial functions
   

## How to connect the splines? 

Knots!

-   Number of knots important
-   Defines flexibility of the line
-   Penalized smoothing spline
    -   Cost to being non-linear
    -   Avoid over-fitting

## In `R`

```{r whale, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
Fin <- read_csv("NewFinLN.csv")
Fin <- Fin[,c('LE','m4','NewStatus','SelP')]
Fin <- na.omit(Fin)       #omit NA's
```

```{r gam, message=FALSE,warning=FALSE, echo=TRUE}
library(mgcv)
fin <- gam(m4~s(LE,bs="ts")+s(SelP,bs="ts")+NewStatus,data=Fin)
summary(fin)
```

## Visual

```{r plotgam, echo=TRUE}
par(mfrow=c(1,2))
plot(fin)
```

## Diagnose

```{r check, echo=TRUE}
par(mfrow=c(2,2))
gam.check(fin)
```

## Diagnostic plots

```{r plots}
resid <- residuals(fin, type = "deviance")
linpred <- napredict(fin$na.action, fin$linear.predictors)
observed.y <- napredict(fin$na.action, fin$y)
par(mfrow=c(2,2))
qq.gam(fin, rep = 0, level = 0.9, type = "deviance", rl.col = 2, 
       rep.col = "gray80")
hist(resid, xlab = "Residuals", main = "Histogram of residuals")
plot(linpred, resid, main = "Resids vs. linear pred.", 
     xlab = "linear predictor", ylab = "residuals")
plot(fitted(fin), observed.y, xlab = "Fitted Values", 
     ylab = "Response", main = "Response vs. Fitted Values")

```

## More on GAMs

Change the family
Add random effects (GAMM)

Great online tutorial at: [https://noamross.github.io/gams-in-r-course/](https://noamross.github.io/gams-in-r-course/)

## Generalized Estimating Equations (GEEs)

Method for modelling longitudinal or clustered data

Why not a mixed model?

. . . 

-   GEE models population average, MM are subject specific models
-   GEE doesn't use a likelihood so can be easier to compute
-   Can specify the correlation structure

## Why not a GEE?

-   Doesn't handle complex study designs well
-   Model not suitable for simulation
-   A lot of user definition required

. . . 

`geepack` in `R` for implementation

## Ridge Regression

Addresses two issues:

-   Number of parameters exceeds the number of observations
-   Predictor variables are highly correlated (collinearity)

. . . 

Works to decrease variance, which increases bias

-   Bias is towards the null
-   Shrinkage methods - shrinks the regression coefficients towards zero

## Least Abosolute Shrinkage Selection Operator (lasso)

Same idea as ridge regression

-   Both are penalized regression
-   Both are shrinkage methods

Use different penalties

-   Lasso can zero out coefficients
-   Ridge is "easier" to implement, faster to compute

## Multivariate regression

**Multivariate $\ne$ Multi-variable**

Multiple response variables in one model

-   Model the correlation between response variables
-   Same predictors for each, but different relationships
-   Don't need special `R` packages

## Multivariate regression in `R`
```{r mult, echo=TRUE}
data("airquality")
ozone <- airquality[complete.cases(airquality),]
m1    <- lm(cbind(Ozone,Solar.R)~Temp+Wind,data=ozone)
summary(m1)
```

## Classification and Regression Trees (CART)

Common approach to classification

-   Sequential, binary splits of the data
-   Classification tree: Categorical data
-   Regression tree: Continuous data

. . . 

Three main elements 

-   Rules for splitting nodes based on a single variable
-   Stopping rules for when no further splits are possible
-   Prediction for target variable in each terminal node

## Advantages of CART

-   Non-parametric
-   Not significantly impacted by outliers
-   Same variables can occur in multiple places
-   Can deal with complex inter-dependencies
-   Can be used in conjunction with other prediction methods

## Random Forest

Generate bootstrapped trees

-   From random sample of the variables
-   With replacement

Classify a case for each “tree” in the “forest”

Combine results to decide outcome

-   Average for a regression tree
-   Majority vote for a classification tree

## Always more

Regression is an active field of research

Some models can be re-figured as regression

-   Cox proportional hazards model

Plenty I've never done

-   Elastic net
-   Support vector regression

. . . 

The tool is out there
