---
title: "Linear Regression Assessment, Diagnostics & Interpretation"
author: ""
date: "June 2025"
output:
  xaringan::moon_reader:
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '4:3'
---

```{r, setup-lm1, include=FALSE}
require(mosaic)   # Load additional packages here 
require(ggformula)
require(tidyverse)
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 7, 
                      fig.height = 3,
                      tidy = FALSE,
                      fig.align = 'center', 
                      message = FALSE, 
                      warning = FALSE,
                      error = TRUE,
                      out.width = '80%', 
                      dpi = 300)
theme_set(theme_minimal(base_size = 22))

set.seed(4)
my_data <- data_frame(predictor = rnorm(15, mean = 5, sd = 2),
                      pred1 = predictor,
                      pred2 = gl(2, 8, length = 15, labels = c("A", "B")),
                 response = 1.3 - 0.7* predictor + rnorm(15, 0, 2))


```

# Model Diagnostics and Assessment

- Does the model fit data well?
- *Should* we have fit a line -- Is model appropriate for data?
- Are predictors *really* associated with response?

---

## Regression Conditions

#L
--
inearity

#I
--
ndependence

#N
--
ormality of residuals

#E
--
qual variance

---

# Flowers

Iris data set in `R`

--
- Is there a relationship between the length of an iris petal, its width and its species?

```{r, echo = TRUE}
# Load the data
library(datasets)

# Look at the data
glimpse(iris)
```
---

# Our Model

```{r, echo=TRUE}
# Use informative model names!
flower <- lm(Petal.Length ~ Petal.Width + Species, 
          data = iris)

# and data should be named informatively, too!
my_iris <- iris |>
  mutate(preds = predict(flower),
         resids = resid(flower))
```

---

# Linearity: Data Exploration

### No trend, Linear trend: **OK**

```{r, echo=TRUE}
# Visualize the data
ggplot(iris,aes(Petal.Width,Petal.Length,color=Species)) +
  geom_point()+
  labs(title = "Length and Width of Iris Petals",
       x = "Petal Width (cm)",
       y = "Petal Length (cm)",
       color = "Species")
```

---

# Linearity: Residual Plots
### No trends: **OK**

```{r, echo=TRUE}
gf_point(resids ~ preds,
         data = my_iris)
```

---

# Your turn!

Using the multiple regression model from the ultra-processed foods data set:
   - Extract the residuals and fitted values
   - Plot the response variable against one of the explanatory variables
   - Plot the residuals against the fitted values
   


---
## Independence of Residuals
### Ponder sort order; then ACF plot
### Logic and knowledge of data set

```{r, echo=TRUE}
acf(my_iris$resids, main="Residual ACF plot")
```

---

# Your turn!

Using the same multiple regression model:
   - Create an acf plot


---

## Normality of the Residuals: Histogram
### (Be quite generous)

```{r}
gf_histogram(~resids, data = my_iris)
```

---
## Normality of the Residuals: Q-Q plot
### (Be quite generous)

```{r}
gf_qq(~resids, data = my_iris) |>
  gf_qqline()
```

---

# Your turn!

Using the multiple regression model from the ultra-processed foods data set:
   - Plot a histogram of the residuals
   - Generate a QQ plot
   

---

## Constant Residual Variance
### Point cloud should fit well in a rectangle (no trumpet or curve)
### True for residual vs. fitted *and* residual vs. any predictor

```{r, message = FALSE, warning = FALSE}
gf_point(resids ~ preds, data = my_iris, title = 'OK', size = 4) |>
  gf_hline(yintercept = 3, color = 'darkred', linetype = 'dashed', size = 3) |>
  gf_hline(yintercept = -3, color = 'darkred', linetype = 'dashed', size = 3) |>
  gf_hline(yintercept = 0, color = 'black', linetype = 'dashed', size = 2) |>
  gf_labs(x="Fitted", y="Residuals")
```

---

# Your turn!

Using the multiple regression model from the ultra-processed foods data set:
   - Plot the residuals against the fitted values
   - Add dashed lines 3 standard deviations from the mean
   

---

# *Any* LINE Violation -> Danger!
## Conclusions can not be trusted

- slope estimates **incorrect**
- CIs and p-values **too small**
- poor prediction accuracy

--
- Most robust to linearity and normality
- Independence and equal variance the kickers

---

# $R^{2}$
### Simple measure of fit of model to data
### NOT *goodness* of model or *appropriateness*

--
- Ranges from 0 - 1
  - 0: no trend
  - 1: perfect fit
- Percentage of variance explained 


$$R^2 =1 - \frac{\sum_{i=1}^{n} (y_i - \hat{y}_i)^2} {\sum_{i=1}^{n} (y_i - \bar{y})^2}$$
---

# Model Output

```{r echo=TRUE}
summary(flower)
```
---

# Interpretation

-  `(Intercept)`: An *Iris setosa* with a petal width of zero has a petal length of 1.21 cm

--
-  `Petal.Width` slope: For each cm increase in petal width, petal length also increases by a cm

--
-  `Speciesverisicolor` slope: The length of *Iris versicolor* petals are, on average 1.7 cm longer than *I. setosa*

--
-  `Speciesvirginica` slope: The length of *Iris virginica* petals are, on average 2.3 cm longer than *I. setosa*

--
-  The length of an iris flower petal is positively associated with its width ($\beta_{1}$ = 1.02, SE = 0.15), and is different between species. 

$$Length_{i} = 1.21 + 1.02Petal.Width_{i} + 1.7versicolor_{i} + 2.3virginica_{i} + \epsilon_{i}$$

---

# Your turn!

Using the multiple regression model from the ultra-processed foods data set:
  -  Interpret your model results
     - Is there an associated between the explanatory variables and the response?
     - What is the nature of the association?
     - How would you communicate your results to another scientist? To the general public?

